Mixture Models for Single-Cell Assays with Applications to Vaccine Studies

### 2013-06-04 MIMOSA.tex
Added calculation of \tilde{z}_i to the supplementary material
Removed some references. Almost there!

\section{Discussion}
\label{s:discussion}
The development of effective statistical methods to detect differences in gene or protein expression at the single-cell level is becoming increasingly important as single-cell assays become more routine and  sequencing at the single-cell level eventually becomes practical~\citep{Ramskold:2012gj}. Current approaches, such as the t-test, $\chi^2$ test and Fisher's exact test are for the most part simplistic, and resulting inference can be quite sub-optimal, especially when the cell counts are small. Most importantly, these methods do not share information across samples, resulting in less power to detect true differences than empirical-Bayes and hierarchical modeling approaches, which are widely applied in the microarray literature~\citep{Kendziorski:2003uw,Newton2004,Smyth:2005iy,Gottardo:2006}. In addition, most of these methods are univariate in nature and inappropriate for high--dimensional, next--generation single-cell assays.

@@

Detecting differences in poly-functional cell populations is important in immunology, since it allows the identification of more homogeneous cell populations~\citep{Milush:2009bz}.
In the context of HIV, poly-functional cell populations have been shown to be correlated with long-term disease non-progression, while in the context of vaccination studies poly--functional responses have been correlated with protection from disease~\citep{Betts:2006dw,Darrah:2007ih,Precopio:2007ht}.
In the ICS data used here, the stimulation is expected to only increase the number of antigen specific cells. The univariate MIMOSA model allows us to constrain the alternative hypothesis to the case $p^{(s)} > p^{(u)}$, where the proportion of cells in the stimulated sample is strictly greater than the proportion of cells in the unstimulated sample, and is ideal for this situation.
Hence, if a specific cell subset expressing multiple biomarkers is differentially expressed, then differential expression based on the marginal cell counts should also be detectable.
As such, identifying poly-functional cytokine profiles from ICS data could be done in an iterative way.
First, univariate tests on marginal populations are performed, then specific cell subsets expressing the positive biomarkers detected are tested.
However, this iterative (univariate) approach might not be satisfactory due to the large number of combinations that need to be tested. A multivariate approach might be preferable.
In that case,  as others have pointed out, in order to have the most power to detect a true difference, the statistical test should only take into account the cytokine combinations of interest~\citep{Nason:2006dx}.

@@

\begin{figure} %  figure placement: here, top, bottom, or page
   \centering
%   \includegraphics[width=0.75\columnwidth]{FIG1.eps}
   \caption{Performance of MIMOSA (EM and MCMC implementations, one-sided model) and competing methods on ICS data from the example flow cytometry data set. Sensitivity and specificity (ROC analysis) as well as observed and nominal false discovery rates for positivity calls from CD4+ T-cells stimulated with A-B) ENV-1-PTEG and expressing IFN-$\gamma$ or C-D) ENV-1-PTEG and expressing IL2. E-F) ENV-1-PTEG and expressing IFN-$\gamma$ or IL2. ROC and FDR plots of other cytokine combinations can be found in Supplementary Figure 1.This figure appears in color in the electronic version of this article.}
\label{fig:HVTN065}
\end{figure}

>>>>

\section{Discussion}
\label{s:discussion}
The development of effective statistical methods to detect differences in gene or protein expression at the single-cell level is becoming increasingly important as single-cell assays become more routine and  sequencing at the single-cell level eventually becomes practical~\citep{Ramskold:2012gj}. Current approaches, such as the t-test, $\chi^2$ test and Fisher's exact test are for the most part simplistic, and resulting inference can be quite sub-optimal, especially when the cell counts are small. Most importantly, these methods do not share information across samples, resulting in less power to detect true differences than empirical-Bayes and hierarchical modeling approaches, which are widely applied in the microarray literature~\citep[e.g.][]{Newton2004,Gottardo:2006}. In addition, most of these methods are univariate in nature and inappropriate for high--dimensional, next--generation single-cell assays.

@@

Detecting differences in poly-functional cell populations is important in immunology, since it allows the identification of more homogeneous cell populations~\citep{Milush:2009bz}.
In the context of HIV, poly-functional cell populations have been shown to be correlated with long-term disease non-progression, while in the context of vaccination studies poly--functional responses have been correlated with protection from disease~\citep[][e.g.]{Betts:2006dw,Precopio:2007ht}.
In the ICS data used here, the stimulation is expected to only increase the number of antigen specific cells. The univariate MIMOSA model allows us to constrain the alternative hypothesis to the case $p^{(s)} > p^{(u)}$, where the proportion of cells in the stimulated sample is strictly greater than the proportion of cells in the unstimulated sample, and is ideal for this situation.
Hence, if a specific cell subset expressing multiple biomarkers is differentially expressed, then differential expression based on the marginal cell counts should also be detectable.
As such, identifying poly-functional cytokine profiles from ICS data could be done in an iterative way.
First, univariate tests on marginal populations are performed, then specific cell subsets expressing the positive biomarkers detected are tested.
However, this iterative (univariate) approach might not be satisfactory due to the large number of combinations that need to be tested. A multivariate approach might be preferable.
In that case,  as others have pointed out, in order to have the most power to detect a true difference, the statistical test should only take into account the cytokine combinations of interest~\citep{Nason:2006dx}.

@@

\begin{figure} %  figure placement: here, top, bottom, or page
   \centering
   \includegraphics[width=0.75\columnwidth]{FIG1.eps}
   \caption{Performance of MIMOSA (EM and MCMC implementations, one-sided model) and competing methods on ICS data from the example flow cytometry data set. Sensitivity and specificity (ROC analysis) as well as observed and nominal false discovery rates for positivity calls from CD4+ T-cells stimulated with A-B) ENV-1-PTEG and expressing IFN-$\gamma$ or C-D) ENV-1-PTEG and expressing IL2. E-F) ENV-1-PTEG and expressing IFN-$\gamma$ or IL2. ROC and FDR plots of other cytokine combinations can be found in Supplementary Figure 1.This figure appears in color in the electronic version of this article.}
\label{fig:HVTN065}
\end{figure}
